---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?


```{r}
steps_per_day <- aggregate(steps ~ date, data = activity, sum)
mean_steps_per_day <- mean(steps_per_day$steps)
median_steps_per_day <- median(steps_per_day$steps)
hist(steps_per_day$steps,
     main = "Histogram using data with NAs",
     col = "blue",
     xlab = "Total steps per day")
abline(v=mean_steps_per_day, col="red")
```

The mean is `r mean_steps_per_day` 
and median is `r median_steps_per_day` of the total number of steps taken per day.

## What is the average daily activity pattern?

```{r}
average_steps <- aggregate(steps ~ interval, data = activity, mean)
max_steps <- max(average_steps$steps)
max_steps_interval <- average_steps[which(average_steps$steps == max_steps),]$interval
plot(average_steps, type="l",
     main = "Average activity in 5-minute intervals per day",
     ylab = "Average Steps",
     xlab = "Minutes")
abline(v=max_steps_interval, col="red")
```

The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps is `r max_steps_interval` with `r max_steps` steps.

## Imputing missing values

Step values with missing data (NAs) have been replaced by the average value for the corresponding interval.

```{r}
total_nas <- sum(is.na(activity$steps))
new_dataset <- activity
for (row in seq(nrow(new_dataset))) {
  if (is.na(new_dataset[row,]$steps)) {
    new_dataset[row,]$steps <- 
      average_steps[which(average_steps$interval == 
                            new_dataset[row,]$interval),]$steps
  }
}
new_steps_per_day <- aggregate(steps ~ date, data = new_dataset, sum)
new_mean_steps_per_day <- mean(new_steps_per_day$steps)
new_median_steps_per_day <- median(new_steps_per_day$steps)
new_average_steps <- aggregate(steps ~ interval, data = new_dataset, mean)
new_max_steps <- max(new_average_steps$steps)
hist(new_steps_per_day$steps,
     main = "Histogram using data with NAs replaced by average values",
     col = "blue",
     xlab = "Total steps per day")
abline(v=new_mean_steps_per_day, col="red")
```

The  total number of missing values in the dataset is `r total_nas` out of `r nrow(activity)` entries, or `r as.integer(100*total_nas/nrow(activity))`%.

The new mean is `r new_mean_steps_per_day` steps per day (same as the original `r mean_steps_per_day` steps) while the new median has risen a step to `r new_median_steps_per_day` steps (the same as the mean) from `r median_steps_per_day` steps. 


The maximum number of steps has remained constant with `r new_max_steps` steps counted for the imputed and `r max_steps` steps for the original data, indicating there were no or few NAs in the most active days.

The days with lots of NA values will obviously have higher step counts, but the average and maximum values are unchanged. The main thing that has changed is the frequency, with the count on the vertical axis rising to 35 from 25.


## Are there differences in activity patterns between weekdays and weekends?

```{r}
new_dataset$daytype <- weekdays(new_dataset$date)
for (row in seq(nrow(new_dataset))) {
  if (new_dataset[row,]$daytype %in% c("Saturday", "Sunday")) {
    new_dataset[row,]$daytype <-  "weekend"
  }
  else {
    new_dataset[row,]$daytype <-  "weekday"  
  }
}
weekend_average_steps <- aggregate(steps ~ interval, 
                                   data = new_dataset[which(new_dataset$daytype == "weekend"),], 
                                   mean)
weekday_average_steps <- aggregate(steps ~ interval, 
                                   data = new_dataset[which(new_dataset$daytype == "weekday"),], 
                                   mean)
par(mfrow=c(2,1))
plot(weekend_average_steps, type="l",
     main="Weekend")
plot(weekday_average_steps, type="l",
     main="Weekday")
```



